---
title: "Week 3: 3 August 2018 --- Data visualisation and R basics"
output: 
  html_document:
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height = 3)
```

We read through the second half of the [third chapter](http://r4ds.had.co.nz/data-visualisation.html), and the [fourth](http://r4ds.had.co.nz/workflow-basics.html). I was chairing the session. Here is a collection of topics that were discussed and my code for reference.

```{r, message=FALSE, echo=FALSE, results=FALSE}
library(tidyverse)
```

## So many categories, so few colours
In section 3.3, ggplot advises against using the **size** of points in a scatterplot to represent a **discrete, categorial** variable (in this case, the class of vehicle). 
    
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

If told to use **shape** to represent the third variable, ggplot uses six shapes at most and left the seventh class (SUVs) unplotted in this case.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```
    
A similar warning does not appear if one uses **colour**, which surprises me. In my eyes, the plot below with different colours for the seven categories looks confusing. **How many colours can the human eye comfortably distinguish?** I can do four at most.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

Sometimes, one should consider using faceting instead of putting everything in a single plot. Note how I used `show.legend = FALSE` to hide the legend, as it would have been redundant given we now have the individual headers.
```{r, fig.height = 5, fig.width = 7}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class), show.legend = FALSE) +
  facet_wrap(~class)
```

Facetting the data this way allows us to clearly visualise individual trends by class. Here, the overall negative correlation between the two variables (on x- and y-axes) holds true even when the data was partitioned by a third variable. However, this would **not** always be the case, and one can bump into a case of [**Simpson’s paradox**](https://www.data-to-viz.com/caveat/simpson.html). 

This discussion is related to question 4 in section 3.5.1. Also check out this [blogpost](https://blog.graphiq.com/finding-the-right-color-palettes-for-data-visualizations-fcd4e707a283), and [Rbrewer](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/) colours which feature colour pallettes suitable for colour blind individuals.

## Putting the layer(s) in "the layered grammar of graphics"

To come

```{r}
p <- ggplot(data = mpg, mapping = aes(x = displ)) + 
  geom_tile(aes(x = 3.65, y = 16, width = 1, height = 6),
            alpha = 0.5, colour = NA, fill = "gold") +
  geom_point(mapping = aes(y = cyl)) +
  labs(y = "cyl (black), cty (red) and hwy (blue)")
```

```{r}
p + geom_point(mapping = aes(y = cty), colour = "red") +
  geom_point(mapping = aes(y = hwy), colour = "blue") +
  ggtitle("blue on top of red")
```

```{r}
p + geom_point(mapping = aes(y = hwy), colour = "blue") +
  geom_point(mapping = aes(y = cty), colour = "red") +
  ggtitle("red on top of blue")
```

## The peculiar aes mapping known as "group = 1"

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = ..prop..))
```

```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, y = ..prop.., group = 1))
```

```{r}
# Multiple groups with one aesthetic
sample_data <- nlme::Oxboys %>% 
  filter(Subject %in% c(10,19,8,25,22))

h <- ggplot(data = sample_data, 
            mapping = aes(x = age, y = height)) + 
  geom_point(mapping = aes(colour = Subject)) 
h
```

```{r, message=FALSE}
# plots a line of best fit for each group
h + geom_smooth(mapping = aes(group = Subject))
```

```{r, message = FALSE}
## fits a single line of best fit through all points
h + geom_smooth(mapping = aes(group = 1))
```

## Statistical transformation = typical confusion
to come

## Reversing the assignment `<-` sign: you can do that?!
This is how you **assign** a variable in R.
```{r}
x <- 3 - 1
x
```

But did you know you can also do it this way? One of the perks in R!
```{r}
2 + 5 -> y
y
```

Using the assignment operator in a reversed fashion is especially useful to capture the output **following a long list of successive operations**. It seems more intuitive to have it at the end rather than the start.

```{r, eval = FALSE}
#this is just a toy example / pseudocode
#we'll learn more about these functions in Chapter 5!
data %>%
  filter(flower_colour %in% c("red", "green")) %>%
  mutate(area = pi*radius^2) %>%
  group_by(petal_shape) %>%
  summarise(mean.area = mean(area)) -> data_summary
```

Above, I introduced **pipes** `%>%`, which would be covered in [Chapter 18](http://r4ds.had.co.nz/pipes.html). It serves as a means for one to feed the output of each step to the next. Without pipes, one could alternatively write the code as:

```{r, eval=FALSE}
#so much redundancy and repetition, prone to typos
data1 <- filter(data, flower_colour %in% c("red", "green"))
data2 <- mutate(data1, area = pi*radius^2) %>%
data3 <- group_by(daTa2, petal_shape) %>%
data_summary <- summarise(data3, mean.area = mean(area)) 
```

The intermediate steps assign variables to memory, which might be problematic if the initial dataset was huge. Such code is not only more difficult to read, but is also more **prone to mistakes**. In fact, can you find the typo in the code above? Now, try to imagine debugging code with long variable names like `HereIsAnUnrealisticallyLongVariableNameJustToProveMyPoint`.




