---
title: 'Week 4: 14 September 2018 --- Data Transformation'
output: 
  html_document:
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height = 3)
```

We read through the [fifth chapter](http://r4ds.had.co.nz/data-visualisation.html) of the book. It was about data transformation with **dplyr**, using its primary functions `filter()`, `arrange()`, `select()`, `mutate()`, `summarise()` and `groupby()`. Personally I am already familiar these, so here I focus on eccentric trivia and tidbits I liked in the chapter.

```{r, message=FALSE}
#import the libraries we need
library(tidyverse)
```

## NA values (Q4 in Exercise 5.2.4)

* Almost any operation with `NA` generally results in an unknown value.

```{r, collapse = TRUE}
NA > 5

NA + 10

10 == NA
```

* Assuming `NA` is a number, the range of values it can take is `{-Inf, Inf}`. Any number raised to the power of zero is 1, therefore the following will not return an unknown value.

```{r}
#note that in the second, it calculates NA ^ 0 first before negative
c( NA ^ 0, -NA ^ 0, (-NA) ^ 0 )
```

* Based on the documentation for `?logical`, logical vectors are coerced to integer vectors in contexts where a numerical value is required, with `TRUE` being mapped to `1L`, and `FALSE` to `0L`:

```{r, collapse = TRUE}
#note that anything that is neither 0 nor 1 does not equate to TRUE/FALSE
c(-Inf, -10, -1, 0, 1, 10, Inf) == TRUE
c(-Inf, -10, -1, 0, 1, 10, Inf) == FALSE
```

This means that `NA` can be either `TRUE` or `FALSE`. We don't know. This leads to such behaviours with the `|` and `&` operators.

```{r, collapse = TRUE}
#only one side needs to be TRUE
NA | TRUE 

#both sides needs to be TRUE, otherwise FALSE
NA & FALSE 

#the answer can still be FALSE if left side is FALSE
NA & TRUE 

#the answer can still be TRUE if left side is TRUE
NA | FALSE 
```

* `NA * 0` does not result in 0, as one might expect.  

```{r}
NA * 0
```

This is because `Inf * 0` evaluates as `NaN` for 'Not a Number'. Take note of the following behaviours.

```{r, collapse = TRUE}
Inf * 0

pi / 0 #finite values divided by 0

0 / 0 
```

* What do you think `Inf - Inf` evaluates to? Check out this [link](https://www.philforhumanity.com/Infinity_Minus_Infinity.html).

## TRUE/FALSE versus truthy/falsy (a minor technicality)

Every value in R has an inherent boolean value, generally known as either **truthy** or **falsey**. This is **different** from being **equal** to `TRUE/FALSE`, which is what the `==` operator tests for!

Remember from above that only `0 == FALSE` and `1 == TRUE`, everything else is neither.

```{r, collapse = TRUE}
x <- c(0, 1, pi, Inf, -1, -pi, -Inf)

x == TRUE

x == FALSE
```

However, **all positive and negative numbers except 0** are inherently **truthy**, which one can see when they are run as a condition of an `if` statement.

```{r, collapse = TRUE}
ifelse(x, 'truthy', 'falsey')
```

Have a go and check how `NA` and `NaN` are assessed. Spoiler: errors incoming!

## A useful idiom I tend to use to do rowwise summaries

Let's say, in this *unrealistic* task, for each row, we want to calculate the mean of values that belong to columns containing the term "time". 

```{r, collapse = TRUE}
#provide a glimpse of the dataset we are working with
library(nycflights13)

colnames(flights)
```

In other words, we want to calculate the mean between `dep_time`, `sched_dep_time`, `arr_time`, `sched_arr_time` and `air_time` **for each row**. Let's take only a subset of the dataset so it's easier to see. I also added a unique index to each row.

```{r}
data_subset <- flights %>% 
  select(month, dep_time, sched_dep_time, arr_time, sched_arr_time, air_time) %>% 
  mutate(unique_id = 1:nrow(flights))

head(data_subset)
```

The function `summarise` only works in a column fashion. To get around the problem, what I do is first `gather` the values from the five "_time" columns into a single column. I then use `group_by` (based on unique identifiers) and `summarise` to calculate the mean, which I rejoin with the original data frame.

```{r, message = FALSE}
data_subset %>% 
  gather("key", "value", contains("time")) %>% 
  group_by(unique_id) %>% 
  summarise(mean_time = mean(value)) %>% 
  full_join(data_subset)
```

For such a task, I used to use `rowwise()` but has been [advised](https://community.rstudio.com/t/dplyr-alternatives-to-rowwise/8071?page=2) that the function is not vectorised and therefore inefficient. There might still be a less clunky way to do what I showed --- please email me if you know!

## Miscellaneous bits

* The following statements are equivalent due to **de Morgan's Law**.

```{r, results = FALSE}
flights %>% filter(arr_delay <= 120, dep_delay <= 120) 
flights %>% filter(!(arr_delay > 120 | dep_delay > 120)) 
```

* Wrapping an assignment in parentheses both prints out the results, and saves them to a variable.

```{r}
(dec25 <- filter(flights, month == 12, day == 25))
```

* Here is a list of verbs/functions I found useful from the chapter:
    * `near` (section 5.2.1): useful when testing two numbers for equality due to finite precision arithmetic.
    * `between` (Q2 in exercise 5.2.4)
    * `everything` (section 5.4): used in conjunction with `select()`
    * `one_of` (Q3 in section 5.4.1)
    * `lead` and `lag` (section 5.5.1)