---
title: "Week 2: 20 July 2018 --- Data visualisation"
output: 
  html_document:
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We read through the first half of the [third chapter](http://r4ds.had.co.nz/data-visualisation.html) about data visualisation using `ggplot2`.

```{r, message=FALSE, echo=FALSE, results=FALSE}
library(tidyverse)
```

## Questions for review (answers in the book)

1. What is an **aesthetic**? Name three examples of aesthetic in ggplot2. (section 3.3) 

2. What are **facets**? 
    * Distinguish between `facet_wrap()` and `facet_grid()`. (section 3.5)

3. What is a **geom**? (section 3.6)
    * What's the difference between **global** and **local** mapping?

4. What is a **stat**? (section 3.7)
    * Differentiate between the stats `"identity"` and `"count"` do in the context of `geom_bar()`.
    * Which is the default?

5. What is a **position adjustment**? (section 3.8)
    * Describe what the position adjustments `"stack"`, `"identity"`, `"dodge"` and `"fill"` do in the context of `geom_bar()`.
    * Which is the default?
    * How does `position = jitter` help with overplotting in scatterplots?

6. What is the default **coordinate system** in ggplot2?

## Questions to ponder (answers in the brain)

1. Distinguish between discrete and continuous data. For these aesthetics, discuss which is suitable for discrete, continuous data, both or neither.
    * size
    * shape
    * colour
    * linetype
    * alpha
    * fill
    
    Consider: discrete data can be categorial or non-categorial.
    
2. With respect to **geom_point** objects, what is the maximum number of dimensions of our data could be simultaneously represented within a **single** ggplot object?
    * Consider we have two axes, various aesthetics and facetting at our disposal.

3. Human eyes are positioned horizontally, giving us a **wider stereoscopic vision**. 
    * Is that why we prefer to have more columns than rows when arranging subplots in a plot (related to question 6 in section 3.5.1)? 
    * Perhaps that's why computer screens were designed to be longer horizontally than vertically? Why did modern smartphones reverse the trend?


## Interesting discussions 

* There are over 30 geoms and 20 stats in ggplot2, and they tell different stories about the same dataset. Check out their official [cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf), as well as the [from Data to Viz](https://www.data-to-viz.com/) website. The latter features a  [section](https://www.data-to-viz.com/caveats.html) about the caveats of commonly used plots, and how one could work around their pitfalls:
    * **Boxplots** summarise a distribution into several numbers: Q1, median, Q3, etc. However, as described [here](https://www.data-to-viz.com/caveat/boxplot.html), we lose information about the sample size and underlying data distribution.
    * When comparing a pair of observations, the classical **barplots with error bars** can be misleading. This [page](https://www.data-to-viz.com/caveat/error_bar.html) describes how outliers and unequal sample sizes between the compared datasets can lead us to make false conclusions.

* Is **mean** or **median** better to summarise data? Would the sample size affect your choice?

* Bar charts pose notable limitations for data visualisation. One point mentioned was that the bars have to be anchored at a specific value, usually zero. Stacked bar charts might be difficult to interepret, therefore try to separate the stacks (use  `position = "dodge"`). Visualisation gets even easier if the bars are ordered by height.

    
## Random code chunks from the book and meeting

Here is a template for any type of plot with ggplot2. The template emphasises the layered **grammar of graphics**, which is what "gg" in ggplot2 stands for. Confession: I only realised this after half a year of using ggplot2.

```{r, eval=FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```

**Every geom has a default stat, and every stat has a default geom.** The following code chunks generate the same plot. If desired, one could change the statistical transformation of a particular geom.

```{r, eval=FALSE}
ggplot(data = diamonds) + 
    geom_bar(mapping = aes(x = cut))

ggplot(data = diamonds) + 
    stat_count(mapping = aes(x = cut))
```

* Conflict between packages can occur if the packages have function(s) of a similar name. An example is the `filter()` function from `dplyr` and `plyr`. Avoid loading too many packages at once. Here is a way to use a function without loading the entire package:

```{r, eval=FALSE}
dplyr::filter()
plyr::filter()
```
