---
title: "Week 3: 3 August 2018 --- Data visualisation and R basics"
output: 
  html_document:
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We read through the second half of the [third chapter](http://r4ds.had.co.nz/data-visualisation.html), and the [fourth](http://r4ds.had.co.nz/workflow-basics.html). I was chairing the session. Here is a collection of topics that were discussed and my code for reference.

```{r, message=FALSE, echo=FALSE, results=FALSE}
library(tidyverse)
```

## So many categories, so few colours
In section 3.3, ggplot advises against using the **size** of points in a scatterplot to represent a **discrete, categorial** variable (in this case, the class of vehicle). 
    
```{r, fig.height = 3, fig.width = 5}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

If told to use **shape** to represent the third variable, ggplot uses six shapes at most and left the seventh class (SUVs) unplotted in this case.
```{r, fig.height = 3, fig.width = 5}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```
    
A similar warning does not appear if one uses **colour**, which surprises me. In my eyes, the plot below with different colours for the seven categories looks confusing. **How many colours can the human eye comfortably distinguish?** I can do four at most.
```{r, fig.height = 3, fig.width = 5}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

Sometimes, one should consider using faceting instead of putting everything in a single plot. Note how I used `show.legend = FALSE` to hide the legend, as it would have been redundant given we now have the individual headers.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class), show.legend = FALSE) +
  facet_wrap(~class)
```

Facetting the data this way allows us to clearly visualise individual trends by class. Here, the overall negative correlation between the two variables (on x- and y-axes) holds true even when the data was partitioned by a third variable. However, this would **not** always be the case, and one can bump into a case of [**Simpson’s paradox**](https://www.data-to-viz.com/caveat/simpson.html). 

This discussion is related to question 4 in section 3.5.1. Also check out this [blogpost](https://blog.graphiq.com/finding-the-right-color-palettes-for-data-visualizations-fcd4e707a283), and [Rbrewer](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/) colours which feature colour pallettes suitable for colour blind individuals.


## The peculiar aesthetic mapping known as "group = 1"
to come

## Statistical transformation
to come

## Reversing the assignment `<-` sign: you can do that?!
to come. I mentioned the concept of piping (`%>%`), which would be covered more in [Chapter 18](http://r4ds.had.co.nz/pipes.html).
